Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2018-11-04 Manuscript/Text/manuscript.tex
Dr. Toth corrections

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   email={lrutter@iastate.edu}   % email address
]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
\author[
   addressref={aff6},
   email={bbonning@ufl.edu}
]{\inits{BB}\fnm{Bryony C.} \snm{Bonning}}
\author[
   addressref={aff2},
   email={dicook@monash.edu}
]{\inits{DC}\fnm{Dianne} \snm{Cook}}
\author[
   addressref={aff3,aff4},
   email={amytoth@iastate.edu}
]{\inits{AT}\fnm{Amy L.} \snm{Toth}}
\author[
   addressref={aff5},
   corref={aff5},
   email={adolezal@illinois.edu}
]{\inits{AD}\fnm{Adam} \snm{Dolezal}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
\parttitle{Background}
Parts of Europe and the United States have witnessed dramatic losses in commercially managed honey bees over the past decade to what is considered an unsustainable extent. The large-scale loss of honey bees has considerable implications for the agricultural economy because honey bees are one of the leading pollinators of numerous crops. Honey bee declines have been associated with several interactive factors. Poor nutrition and viral infection are two environmental stressors that pose heightened dangers to honey bee health. In this study, we used RNA-sequencing to examine how monofloral diets and Israeli acute paralysis virus inoculation influence gene expression patterns in honey bees.

\parttitle{Results}
We found a considerable nutritional response, with almost 2,000 transcripts changing with diet quality. The majority of these genes were over-represented for nutrient signaling (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways). In our experimental conditions, the transcriptomic response to viral infection was fairly limited. We only found 43 transcripts to be differentially expressed, some with known immune functions (argonaute-2), transcriptional regulation, and muscle contraction. We created contrasts to determine if any protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance). A similar number of resistance and tolerance candidate differentially expressed genes were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection. We also compared the virus main effect in our study (polyandrous colonies) to that obtained in a previous study (single-drone colonies) and verified significant overlap in differential expression despite visualization methods showing differences in the noisiness levels between these two datasets.

\parttitle{Conclusions}
Through transcriptional contrasts and functional enrichment analysis, we add to evidence of feedbacks between diet and disease in honey bees. We also show that comparing results derived from polyandrous colonies (which are typically more natural) and single-drone colonies (which usually yield more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently less artificial and less noisy. Altogether, this work underlines the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing studies.

@@

Pollen is a main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}. At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}. At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen \cite{jelly, jelly2}. Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees. In particular, honey bees are confronted with a less diverse selection of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}. Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}. Understanding how undiversified diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.

Viral infection was a comparatively minor problem in honey bees until the last century when the ectoparasitic varroa mite (\textit{Varroa destructor}) spread worldwide \cite{miteSpread}. This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}. More than 20 honey bee viruses have been identified \cite{numVirus}. One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family Dicistroviridae \cite{fam}. IAPV infection causes shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}. IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}. Its role in the rising phenomenon of ``Colony Collapse Disorder'' (in which the majority of worker bees disappear from a hive) remains unclear: It has been implicated in some studies \cite{iapvCCD, iapvCCD2} but not in other studies \cite{ccd1, fam, iapvCCD3}. Nonetheless, it is clear that IAPV reduces colony strength and survival.

@@

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms. For example, diet effects on transcriptomics have been studied in the insect model Drosophila. One recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}. CHK kinase-like protein is believed to respond to DNA damage by arresting the cell cycle. This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake. Another recent Drosophila study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}. No differential expression was found in Drosophila melanogaster, a species known to thrive in all three dietary conditions. However, Drosophila mojavensisa and Drosophila arizonae showed substantial differential expression between the dietary conditions: Genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets. Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.

Likewise, studies on the transcriptomic effects of virus infection have been performed in insect models such as Drosophila and mosquito. These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived of the diverse ways of replication and infection cycle evolved by different viruses. In this sense, an intricate set of pro- and anti-virus host factors are involved in each case \cite{cherryReview, swevers}. For example, dicistroviruses Drosophila C virus (DCV) and cricket paralysis virus (CrPV) infection requires ribosomal proteins RpS6, RpL19 and RACK1, but such proteins are dispensable during infection by the rhabdovirus vesicular stomatitis virus (VSV) in Drosophila. In contrast, the autophagy pathway has proven to be an effective antiviral strategy against VSV, but has no role limiting DCV or other positive-sense RNA viruses \cite{swevers}. For an extensive up-to-date review of known antiviral and proviral factors for specific viruses in Drosophila see \cite{swevers}. Model insect studies have shed light on how honey bees respond to viruses because the function of many genes relies on the characterization of orthologue genes. Nonetheless, recent transcriptomic studies in honey bees have improved our understanding of compatible and incompatible virus-bee interactions. Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g. RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}. Added to these, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken}. This path has been studied recently in more depth using transcriptomics, which found known immune factors but also new ones for which indispensability in antiviral response was determined by silencing \cite{Brutscher}. Similar to Drosophila examples, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees from polyandrous colonies, the Galbraith study examined honey bees from single-drone colonies. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our honey bees may also reflect more realistic environmental and genetic simulations. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

\subsection*{Pathogen response results}

We reanalyzed our previously published dataset with a subset that focuses on diet quality and is more relevant to the current study. We briefly show it again here to inform the RNA-sequencing comparison because we reduced the number of treatments from the original published data (from eight to four) \cite{adamInt} as a means to focus on diet quality effects.

As shown in Figure 1, mortality rates of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05). The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 7.04e-06) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant. We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus. Namely, bees fed rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).

As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titers  based on pairwise comparisons: Bees fed rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03). However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02). Overall, we interpreted these findings to mean that high-quality chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).

\subsection*{Main effect DEG results}

@@

GO analysis of the chestnut-upregulated DEGs revealed the following over-represented categories: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1). GO analysis of the rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).

@@

\subsection*{Pairwise comparison of DEG results}

The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 955 (Supplementary table 8, Additional file 1). Some of the trends observed in the main effect comparisons persisted: The diet level appeared to have greater influence on the number of DEGs than the virus level. Across every pair comparing the chestnut and rockrose levels, regardless of the virus level, the number of chestnut-upregulated DEGs was higher than the number of rockrose-upregulated DEGs (Supplementary table 8 C, D, E, F, Additional file 1). For the pairs in which the diet level was controlled, the virus-exposed treatment showed equal to or more DEGs than the control treatment (Supplementary table 8 A and B, Additional file 1). There were no DEGs between the treatment pair controlling for the chestnut level of the virus effect (Supplementary table 8A, Additional file 1). These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).

\subsection*{Prior study comparison results}

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with new visualization techniques that we recently published in our R package bigPint (https://lrutter.github.io/bigPint).

@@

\subsection*{Tolerance versus resistance results}

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs. We again used parallel coordinate lines superimposed onto side-by-side boxplots to visualize these candidate DEGs. To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns. Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate DEGs (Additional files 15 and 16). However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate DEGs, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group. Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15). Likewise, based on the contrasts we created to obtain the `resistance'' candidate DEGs, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups. We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16). The genes in Cluster 3 may follow the expected pattern the most distinctively (Additional file 16).

>>>>

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   email={lrutter@iastate.edu}   % email address
]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
\author[
   addressref={aff6},
   email={bbonning@ufl.edu}
]{\inits{BB}\fnm{Bryony C.} \snm{Bonning}}
\author[
   addressref={aff2},
   email={dicook@monash.edu}
]{\inits{DC}\fnm{Dianne} \snm{Cook}}
\author[
   addressref={aff3,aff4},
   email={amytoth@iastate.edu}
]{\inits{AT}\fnm{Amy L.} \snm{Toth}}
\author[
   addressref={aff5},
   corref={aff5},
   email={adolezal@illinois.edu}
]{\inits{AD}\fnm{Adam G.} \snm{Dolezal}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
\parttitle{Background}
Parts of Europe and the United States have witnessed dramatic losses in commercially managed honey bees over the past decade to what is considered an unsustainable extent. The large-scale loss of bees has considerable implications for the agricultural economy because bees are one of the leading pollinators of numerous crops. Bee declines have been associated with several interactive factors. Recent studies suggest nutritional and pathogen stress can interactively contribute to bee physiological declines, but the molecular mechanisms underlying interactive effects remain unknown. In this study, we provide insight into this question by using RNA-sequencing to examine how monofloral diets and Israeli acute paralysis virus inoculation influence gene expression patterns in bees.

\parttitle{Results}
We found a considerable nutritional response, with almost 2,000 transcripts changing with diet quality. The majority of these genes were over-represented for nutrient signaling (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways). In our experimental conditions, the transcriptomic response to viral infection was fairly limited. We only found 43 transcripts to be differentially expressed, some with known immune functions (argonaute-2), transcriptional regulation, and muscle contraction. We created contrasts to explore whether protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance). A similar number of resistance and tolerance candidate differentially expressed genes were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection.

\parttitle{Conclusions}
Through transcriptional contrasts and functional enrichment analysis, we contribute to our understanding of the mechanisms underlying feedbacks between nutrition and disease in bees. We also show that comparing results derived from combined analyses across multiple RNA-seq studies may allow researchers to identify transcriptomic patterns in bees that are concurrently less artificial and less noisy. This work underlines the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing studies.

@@

Pollen is a main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}. At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}. At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen \cite{jelly, jelly2}. Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees. In particular, honey bees are confronted with a less diverse selection of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}. Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}. Understanding how low diversity diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.

Viral infection was a comparatively minor problem in honey bees until the last century when the ectoparasitic varroa mite (\textit{Varroa destructor}) spread worldwide \cite{miteSpread}. This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}. More than 20 honey bee viruses have been identified \cite{numVirus}. One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV), a positive-sense RNA virus of the family Dicistroviridae \cite{fam}. IAPV infection causes shivering wings, decreased locomotion, muscle spasms, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}. IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}.

@@

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms. For example, diet effects on transcriptomics have been studied in the insect model Drosophila. One recent transcriptomic study in \textit{Drosophila melanogaster} reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}. This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake. Another recent study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}. \textit{Drosophila mojavensis} and \textit{Drosophila arizonae} showed substantial differential expression between the dietary conditions: genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets. Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.

There have been numerous studies on the transcriptomic effects of virus infection in model organisms like fruit flies and mosquitoes that can provide a useful framework for interpreting virus responses in honey bees. These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived from the diverse methods of replication and infection cycle evolved by different viruses. An intricate set of pro- and anti-virus host factors such as ribosomal proteins and autophagy pathways are involved, but the response depends on the virus species, as has been elucidated in Drosophila \cite{cherryReview, swevers}. Model insect studies can inform studies of honey bee virus response, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms. Recent transcriptomic studies in honey bees have improved our understanding of putative molecular mechanisms of honey bee-virus interactions.

Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g. RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}. In addition, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken, Brutscher}. As in Drosophila, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees from naturally mated colonies, the Galbraith study examined honey bees from single-drone colonies. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used naturally mated honey bees from 15 different colonies. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally mated honey bees may also reflect more realistic environmental and genetic simulations. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We wish to explore how the gene expression effects of IAPV inoculation compare between these two studies that used such different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

\subsection*{Mortality and virus titers}

We reanalyzed our previously published dataset with a subset that focuses on diet quality and is more relevant to the current study. We show the data subset here to inform the RNA-sequencing comparison because we reduced the number of treatments from the original published data (from eight to four) \cite{adamInt} as a means to focus on diet quality effects.

As shown in Figure 1, mortality rates of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 54; F = 10.03; p $<$ 2.34e-05). The effect of virus treatment (mixed model ANOVA, df = 1, 54; F = 24.73; p $<$ 7.04e-06) and diet treatment (mixed model ANOVA, df = 1, 54; F = 5.32; p $<$ 2.49e-02) were significant, but the interaction between the two factors (mixed model ANOVA, df = 1, 54; F = 4.72e-02, p = 8.29e-01) was not significant. We compared mortality levels based on pairwise comparisons: For a given diet, honey bees exposed to the virus showed significantly higher mortality rate than honey bees not exposed to the virus. Bees fed rockrose pollen had significantly elevated mortality with virus infection compared to uninfected controls (Benjamini-Hochberg, p $<$ 1.53e-03), and bees fed chestnut pollen similarly had significantly elevated mortality with virus infection compared to controls (Benjamini-Hochberg, p $<$ 3.12e-03) (Figure 1).

As shown in Figure 2, IAPV titers of honey bees 72 hours post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titers  based on pairwise comparisons: Bees fed rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03). However, bees fed chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02). Overall, we interpreted these findings to mean that high-quality chestnut pollen could partially ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality rockrose pollen could not (Figure 2).

\subsection*{Transcriptomic responses to virus infection and diet}

@@

GO analysis of the chestnut-upregulated DEGs revealed the following over-represented biological functions: Wnt signaling, hippo signaling, and dorso-ventral axis formation, as well as pathways related to circadian rhythm, mRNA surveillance, insulin resistance, inositol phosphate metabolism, FoxO signaling, ECM-receptor interaction, phototransduction, Notch signaling, JaK-STAT signaling, MAPK signaling, and carbon metabolism (Supplementary table 2, Additional file 1). GO analysis of the rockrose DEGs revealed pathways related to terpenoid backbone biosynthesis, homologous recombination, SNARE interactions in vesicular transport, aminoacyl-tRNA biosynthesis, Fanconi anemia, and pyrimidine metabolism (Supplementary table 3, Additional file 1).

@@

The number of DEGs across the six treatment pairings between the diet and virus factor ranged from 0 to 955 (Supplementary table 8, Additional file 1). Again, diet level appeared to have greater influence on the number of DEGs than the virus level. Across every pair comparing the chestnut and rockrose levels, regardless of the virus level, the number of chestnut-upregulated DEGs was higher than the number of rockrose-upregulated DEGs (Supplementary table 8 C, D, E, F, Additional file 1). Virus-treated bees showed equal to or more upregulated genes relative to controls, under both diet treatments (Supplementary table 8 A and B, Additional file 1). These trends were observed for all three pipelines used (DESeq2, edgeR, and limma).

\subsection*{Transcriptomic data visualization and comparison to a previous study}

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online (https://lrutter.github.io/bigPint) and are in preparation for publication.

@@

\subsection*{Tolerance versus resistance}

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs. Within our 122 ``tolerance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, fructose metabolism, and chitin metabolism). However, we also discovered gene ontologies related to RNA polymerase II transcription, immune response, and regulation of response to reactive oxygen species (Figure 5A). Within our 125 ``resistance'' gene ontologies, we found functions related to metabolism (such as carbohydrate metabolism, chitin metabolism, oligosaccharide biosynthesis, and general metabolism) (Figure 5B).

To visually explore gene expression patterns related to tolerance and resistance, we used hierarchical clustering to separate candidate DEGs into common patterns, and then visualized these clusters using parallel coordinate lines superimposed onto side-by-side boxplots. To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns. Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate DEGs (Additional files 15 and 16). However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate DEGs, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group. Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15). Likewise, based on the contrasts we created to obtain the `resistance'' candidate DEGs, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups. We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16). The genes in Cluster 3 may follow the expected pattern the most distinctively (Additional file 16).
